Extended Data

### 2022-03-29 doc/manuscript/Dow_manuscript_Nature.Rmd
proofreading edits

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020;<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

>>>>

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--@stuble_plant_2021;@delgado_differences_2020-->] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[<!--@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;-->@jeong_phenology_2011;@finzi_carbon_2020<!--@stuble_plant_2021-->] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@crabbe_extreme_2016;@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@savage_coordination_2021;@delpierre_temperate_2016], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting projections from global C cycle models[@richardson_terrestrial_2012;@keenan_net_2014], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity enough to strengthen the long-term CO~2~ sink of temperate deciduous forests.**

@@

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.
